---
title: "VehiclePopRatio_10Counties"
---

# Summary

* This script processes a dataset that was extracted from the Census FactFinder at the Census block group level. 

* The goal is to generate a map showing the percentage of the population  that lacks access to a vehicle. The extent is Philadelphia plus 9 surrounding counties.

* Output: Produces a feature class at the block group level, exported as a Shp.

* Last modified: 11/07/2019

* For more documentation, see https://docs.google.com/document/d/121wuW2rLnx1hJ-_4_Ew7crX0tLyWc8PaEUwvSNwggeU/edit?usp=sharing


# Init

Initialize the paths and names. 

```{r init}

library(sf)
library(ggplot2)
library(writexl)

# Paths
home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data3rdPass/"
tableSourceFolder <- "RawData/RawXls/VehiclesAvailable_10Counties/"
boundarySourceFolder <- "RelevantBoundaries/"
outFolder <- "DisplayReadyData/Shp/"

# Dataset names
origSet <- "ACS_17_5YR_B25044"  #  Tenure by Vehicles Available
geoLayerName <- "10Counties_Blockgroups_Proj"
outputLayerName <- "ZeroVehicleRatio"
```

## Get the vehicle data

```{r get_vehicles}

fileToRead <- paste(home, tableSourceFolder, origSet, "_with_ann.csv", sep="")
fileToRead
acsData <- read.csv(file=fileToRead, header=TRUE, sep=",")
acsData <- acsData[-1, ]

# Keep only the columns of relevance and rename field: 
# GeoID, total number of occupied housing units (OHU), Owner with 0 vehicles,
# Renter with 0 vehicles)
acsData <- acsData[c("GEO.id2", "HD01_VD01", "HD01_VD03", "HD01_VD10")]

# Convert them to the right format (from the original "factor" type)
acsData[,2:4] <- sapply(acsData[,2:4], as.character)
acsData[,2:4] <- sapply(acsData[,2:4], as.numeric)

# Add all OHU with 0 vehicles available (owner + renter), and divide by the total number of OHU
acsData["Total_ZeroVehicule"] <- (acsData$HD01_VD03 + acsData$HD01_VD10) 
acsData["ZeroVehicule_Pct"] <- round((acsData$Total_ZeroVehicule / acsData$HD01_VD01) * 100, digits = 2)
  
colnames(acsData)[colnames(acsData)=="HD01_VD01"] <- "Total_OHU"

# Remove the rows where the total OHU is 0
acsData <- subset(acsData, acsData$Total_OHU > 0 ) 

# Keep only the columns of relevance and reorder the columns
acsData <- acsData[c("GEO.id2","Total_ZeroVehicule", "Total_OHU",  "ZeroVehicule_Pct")]

# Choose short column names for shapefile exports
names(acsData) <- c("GEO.id2","0Vehicle", "TotalOHU",  "0VehPct")

# Sort by descending order (just to review the data)
#acsData[order(acsData$Total_OHU),]
#acsData[order(-acsData$ZeroVehicule_Pct),]

acsData


```

## Export to Excel (Just to do some manual data checks, if desired)

```{r export_to_excel}

fileToWrite <- paste(home, outFolder,outputLayerName, ".xlsx", sep="")
#write_xlsx(x = acsData, path = fileToWrite, col_names = TRUE)

```

## Load PHL geo layer (block group boundaries)

```{r load_geo_layer}

# Read the shapefile
shpToRead <- paste(home, boundarySourceFolder, geoLayerName,".shp", sep="")

geoLayer = st_read(shpToRead)

# Keep only the columns of relevance
geoLayer <- geoLayer[c("GEOID", "geometry")]

#str(geoLayer)
geoLayer
str(geoLayer)

# Plot it
plot(st_geometry(geoLayer), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```

## Merge geolayer and table data

```{r merge_geo_and_data}

geoLayerWithData <- merge(geoLayer, acsData, by.x="GEOID", by.y="GEO.id2")

# Sort by descending order / check for NA values etc. (just to review the data)
#geoLayerWithData[order(-acsData$ZeroVehicule_Pct),] 
#subset(geoLayerWithData,geoLayerWithData$GEOID=="421019804001") 

geoLayerWithData 

```

## Plot map

For testing purposes.

```{r plot_map,  include=FALSE}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

#ggplot() + geom_sf(data = geoLayerWithData,  aes(fill = ZeroVehicule_Pct)) + theme_bw() + #ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```


## Export to Shp

```{r export_to_shp}

shpToWrite <- paste(home, outFolder,  outputLayerName, ".shp", sep="")
shpToWrite
st_write(geoLayerWithData, shpToWrite, delete_layer=TRUE)

```

