---
title: "MinorityStatus_Countylevel_10Counties"
---

# Summary

* Process the CTPP data: Minority Status at the County level for 10 counties (9 DVRPC counties + New Castle, DE.) 

* Input: Single CSV file extracted from the CTPP site representing 10 Counties to 10 Counties flows. Note that the suburb to suburb flows will be naturally eliminated through the process "Keep only PHL as a residence and PHL as a workplace".

* Output: Produces a feature class at the County level with 6 data colums: (1) where do residents of philadelphia work (All, Minority, White)? and (2) where do workers in philadelphia come from  (All, Minority, White)? Both are merged and exported into the same Shp. The same data is also exported to Excel.

* Last modified: 09/26/2019

# Init

Initialize the paths and names. 

```{r init}

library(zoo)
library(dplyr)
library(writexl) 
library(sf)
library(ggplot2)

home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data2ndPass/"

tableName <- "MinorityStatus_Countylevel_10Counties"

tableSourceFolder <- "RawData/RawXls/improved_tads/"
boundarySourceFolder <- "RelevantBoundaries/"
outFolder <- "DisplayReadyData/"

geoLayerName <- "10Counties"


```

# Part I: Read the table data and clean it up

## Read the raw CSV table

```{r read_csv}

# Read the file
  fileToRead <- paste(home, tableSourceFolder, theme, category, ".csv", sep="")
fileToRead
ctppData <- read.csv(file=fileToRead, header=FALSE, sep="," ,skip=6)

# Remove that last line of the table (which is a note)
ctppData <- head(ctppData,-1)

# Remove the Margin of Error fields
noMoE <- c(1,2,3,5,7)
  
ctppData <- ctppData[noMoE]

# Rename the fields
names(ctppData) <- c("RESIDENCE",	"WORKPLACE",	"Total", "White_not_hisp", "Minority")

#str(ctppData)
ctppData

```

## Clean up ID columns

```{r cleanup_ids}

# Create IDs out of Residence and Workplace strings
#ctppData$RESIDENCE <- gsub('TAD (\\d+), MPO (\\d+)', 'C1200US\\2\\1', ctppData$RESIDENCE)
#ctppData$WORKPLACE <- gsub('TAD (\\d+), MPO (\\d+)', 'C3200US\\2\\1', ctppData$WORKPLACE)

# Fill up empty Residence cells (due to the "cell merge" presentation) with last previous value
ctppData$RESIDENCE[ctppData$RESIDENCE == ""] <- NA
ctppData$RESIDENCE <- na.locf(ctppData$RESIDENCE)

# Make all the IDs "neutral" (not C1 or C3, which expresses Residence vs Workplace)
#ctppData$RESIDENCE <- gsub('^C1', 'C0', ctppData$RESIDENCE)
#ctppData$WORKPLACE <- gsub('^C3', 'C0', ctppData$WORKPLACE)

ctppData

```

## Convert quantitative cells from text columns to numeric

``` {r text_to_numeric}

numCols <- ncol(ctppData)

# Turn all quantitative cells to the numeric type, after having removed the potential comma. 
ctppData[3:numCols]<- lapply(ctppData[3:numCols], function(y) as.numeric(gsub(",", "", y)))

ctppData
```


# Part II: Generate two county-level feature classes

## Keep only PHL as a residence and PHL as a workplace

```{r collapse_PHL}

# Change the name of the dataframe (to match a previous version of this script) 
ctppDataAggreg <- ctppData

# Keep only the rows where the residence is Philadelphia County:
ctppDataPhilaAsResidence <- ctppDataAggreg[ctppDataAggreg$RESIDENCE == "Philadelphia County, Pennsylvania", ]
ctppDataPhilaAsResidence

# Keep only the rows where the workplace is a Philadelphia County:
ctppDataPhilaAsWorkplace <- ctppDataAggreg[ctppDataAggreg$WORKPLACE  == "Philadelphia County, Pennsylvania", ]
ctppDataPhilaAsWorkplace

```

## Get PHL as a residence and PHL as a workplace in the same dataframe

```{r Merge_two}
ctppDataPhilaAsResidenceTotal <- ctppDataPhilaAsResidence[c("RESIDENCE",	"WORKPLACE",	"Total", "White_not_hisp", "Minority")]
names(ctppDataPhilaAsResidenceTotal) <- c("RESIDENCE",	"WORKPLACE",	"TotalWp", "WhiteWp", "MinorityWp")

ctppDataPhilaAsWorkplaceTotal <- ctppDataPhilaAsWorkplace[c("RESIDENCE",	"WORKPLACE",	"Total", "White_not_hisp", "Minority")]
names(ctppDataPhilaAsWorkplaceTotal) <- c("RESIDENCE",	"WORKPLACE",	"TotalRe", "WhiteRe", "MinorityRe")

ctppDataPhilaAsResidenceTotal
ctppDataPhilaAsWorkplaceTotal

TenCountyTotals <- merge(ctppDataPhilaAsResidenceTotal, ctppDataPhilaAsWorkplaceTotal, by.x="WORKPLACE", by.y="RESIDENCE")
TenCountyTotals <- TenCountyTotals[c("WORKPLACE",	"TotalWp", "TotalRe", "WhiteWp", "WhiteRe", "MinorityWp", "MinorityRe")]

names(TenCountyTotals)[names(TenCountyTotals)=="WORKPLACE"] <- "COUNTY"

TenCountyTotals

```

## Load 10 County boundaries shapefile

```{r load_geo_layer}

# Read the "10County" shapefile
shpToRead <- paste(home, boundarySourceFolder, geoLayerName, ".shp", sep="")
geoLayer = st_read(shpToRead)

#str(geoLayer)
geoLayer

# Plot it
plot(st_geometry(geoLayer), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```


## Merge county boundaries and data table

```{r merge_geo_and_data}

# Join the non-PHL side to the Greater TADs.
GeoTenCountyTotals <- merge(geoLayer, TenCountyTotals, by.x="name", by.y="COUNTY")

GeoTenCountyTotals
```


## Plot resulting County-level maps

The field Total_Est is displayed.

```{r plot_map, include=TRUE}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

ggplot() + geom_sf(data = GeoTenCountyTotals,  aes(fill = TotalRe)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

ggplot() + geom_sf(data = GeoTenCountyTotals,  aes(fill = TotalWp)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```

# Part 3: Export
## Export to Shapefile

```{r export_to_shp, include=TRUE}

shpToWrite <- paste(home, outFolder, "MinorityStatus_TenCountyTotals.shp", sep="")
#shpToWrite
st_write(GeoTenCountyTotals, shpToWrite, delete_layer=TRUE)

```


## Export to Excel

```{r export_to_excel}

fileToWrite <- paste(home, outFolder,"MinorityStatus_TenCountyTotals.xlsx", sep="")
write_xlsx(x = TenCountyTotals, path = fileToWrite, col_names = TRUE)

```
