---
title: "MeansOfT_PHL_and_PHL"
---

# Summary

* This script processes the Means of Transportation "PHL2PHL" data.

* Input: Data extracted from the CTPP site and already cleaned up. Excel format. 

* Output: Produces one feature class for each of the 6 "PHL area", with the Residence collapsed. Written to a shapefile.

* Last modified: 8/20/2091


# Init

```{r init}

library(sf)
library(readxl) 
library(ggplot2)

home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data/"

tableSourceFolder <- "GisReadyTables/"
boundarySourceFolder <- "RelevantBoundaries/"
outFolder <- "DisplayReadyData/"

flowName <- "PHL2PHL"

```

# Main: Generate 6 PHL Area-level feature classes

## Read the Excel table

```{r read_excel}

fileToRead <- paste(home, tableSourceFolder, "MeansOfT_", flowName, "_GisReady.xlsx", sep="")
fileToRead
ctppData <- read_xlsx(fileToRead, col_names = TRUE)

# Make all the IDs "neutral" (not C1 or C3, which expresses Residence vs Workplace)
ctppData$RESIDENCE <- gsub('^C1', 'C0', ctppData$RESIDENCE)
ctppData$WORKPLACE <- gsub('^C3', 'C0', ctppData$WORKPLACE)

# Rename some of the columns
names(ctppData) <- gsub(x = names(ctppData), pattern = "_Est", replacement = "")  
colnames(ctppData)[colnames(ctppData)=="Transit_Total"] <- "Tran_Tot"

#str(ctppData)
ctppData

# For testing purposes: sum up all rows whose residence is Center City ("C0200US4219650100000101"):
#test <- ctppData[ctppData$RESIDENCE=="C0200US4219650100000101",]
#sum(test$Total)


```


## Join with TADs and "Larger Areas" crosswalk xls

Note: We have divided Philadelphia into 6 Large Areas

```{r join_to_crosswalk}

# Read the crosswalk Excel file
fileToRead <- paste(home, boundarySourceFolder, "PHL_TADs_and_Areas.xlsx", sep="")
PhlTADsAndAreas <- read_xlsx(fileToRead, col_names = TRUE)

# Join it to the main table for both the Residence and Workplace
ctppDataAreas <- merge(ctppData, PhlTADsAndAreas, by.x="RESIDENCE", by.y="neutral_id", all.x = TRUE)
colnames(ctppDataAreas)[which(names(ctppDataAreas) == "Area")] <- "Res_Area"
ctppDataAreas

ctppDataAreas2 <- merge(ctppDataAreas, PhlTADsAndAreas, by.x="WORKPLACE", by.y="neutral_id", all.x = TRUE)
colnames(ctppDataAreas2)[which(names(ctppDataAreas2) == "Area")] <- "Work_Area"

# Reorder the columns
colsToSort <- c("RESIDENCE", "WORKPLACE", "Res_Area", "Work_Area")
othercols <- setdiff(names(ctppDataAreas2), colsToSort)
  
ctppDataAreas2 <- ctppDataAreas2[c(colsToSort , othercols)]
ctppDataAreas2

# For testing purposes: sum up all rows whose residence is Center City ("Center"):
#test2 <- ctppDataAreas2[ctppDataAreas2$Res_Area=="Center",]
#sum(test2$Total)


```

## Collapse PHL TADs to Areas

```{r collapse_PHL}

# Group by Large Areas on both the Residence and the Workplace sides
ctppDataAggreg <- aggregate(list(ctppDataAreas2[5:14]), by=list(Res_Area = ctppDataAreas2$Res_Area, Work_Area = ctppDataAreas2$Work_Area), FUN=sum, na.rm=TRUE)

ctppDataAggreg

```


## Load PHL Large Area geo layer

```{r load_geo_layer}

# Read the shapefile
layerName <- "PHL_Large_Areas"
shpToRead <- paste(home, boundarySourceFolder, layerName, ".shp", sep="")
geoLayer = st_read(shpToRead)

# Keep only the columns of relevance
geoLayer <- geoLayer[c("Large_Area", "geometry")]

#str(geoLayer)
geoLayer

# Plot it
plot(st_geometry(geoLayer), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```

## Generate feature classes for each residence area subset

```{r merge_geo_and_data}

# Generate subsets for each residence area by looping through every area
PhlAreas <- unique(ctppDataAggreg$Res_Area)

# Loop through every residence area
for(a in PhlAreas){
  # Generate the subset for that area
  ctppDataSubset <- ctppDataAggreg[ctppDataAggreg$Res_Area == a,]

  # Compute new percentage columns for every quantitative field
  for(i in colnames(ctppDataSubset[3:12])){
    ctppDataSubset[[paste0(i, "Pct")]] <- round((ctppDataSubset[[i]]/sum(ctppDataSubset[[i]]))*100, digits = 2)
  }
  
  # Merge geolayer with the subset table data
  geoLayerAreas <- merge(geoLayer, ctppDataSubset, by.x="Large_Area", by.y="Work_Area")
  geoLayerAreas

  colnames(geoLayerAreas)[colnames(geoLayerAreas)=="Large_Area"] <- "Workplace"
  colnames(geoLayerAreas)[colnames(geoLayerAreas)=="Res_Area"] <- "Residence"
  
    
  # Export to Shp
  shpToWrite <- paste(home, outFolder, "MeansOfT_LargeAreas_", flowName, "_", a, "_full.shp", sep="")
  shpToWrite
  st_write(geoLayerAreas, shpToWrite, delete_layer=TRUE)
  }


```

## Plot map

For testing purposes

```{r plot_map,  include=FALSE}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

ggplot() + geom_sf(data = geoLayerAreas,  aes(fill = Total)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```