---
title: "MeansOfT_PHL_and_Suburbs"
---

# Summary

* This script processes the Means of Transportation "Greater2PHL" and "PHL2Greater" data. "Greater" stands for Greater Philadelphia and refers to DVRPC's 9 suburban counties.

* Input: Data extracted from the CTPP site and already cleaned up. Excel format. 

* Output: Produces feature classes at the TAD and County levels, with the Residence collapsed. Written to a Shp.

* Last modified: 8/20/2019


# Init

Initialize the paths and names. 

```{r init}

library(sf)
library(readxl) 
library(ggplot2)

home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine/CtppAnalysis/Data/"

tableSourceFolder <- "GisReadyTables/"
boundarySourceFolder <- "RelevantBoundaries/"
outFolder <- "DisplayReadyData/"

# This script can be used in two modes, to process either PHL to Suburbs data or Suburbs to PHL.
# Mode 1:
#flowName <- "PHL2Greater"
#PhlSide <- "RESIDENCE"

# Mode 2:
flowName <- "Greater2PHL"
PhlSide <- "WORKPLACE"

geoLayerName <- "Greater_TADs"
#geoLayerName <- "Greater_TADs_Centroids"


```

# Part I: Generate the TAD level feature class

## Read the Excel table

These were extracted from the CTPP website.

```{r read_excel}

fileToRead <- paste(home, tableSourceFolder, "MeansOfT_", flowName, "_GisReady.xlsx", sep="")
ctppData <- read_xlsx(fileToRead, col_names = TRUE)

# Make all the IDs "neutral" (not C1 or C3, which expresses Residence vs Workplace)
ctppData$RESIDENCE <- gsub('^C1', 'C0', ctppData$RESIDENCE)
ctppData$WORKPLACE <- gsub('^C3', 'C0', ctppData$WORKPLACE)

# Rename some of the columns
names(ctppData) <- gsub(x = names(ctppData), pattern = "_Est", replacement = "")  
colnames(ctppData)[colnames(ctppData)=="Transit_Total"] <- "Tran_Tot"

#str(ctppData)
ctppData

#ctppData[which(ctppData$WORKPLACE == "C0200US4219650100000201"),]$Bus_Est

```


## Collapse PHL to 1 single location

```{r collapse_PHL}

# Group by the non-PHL side (which results in collapsing the PHL side)
if (PhlSide == "RESIDENCE"){
  ctppDataAggreg <- aggregate(list(ctppData[3:12]), by=list(WORKPLACE = ctppData$WORKPLACE), FUN=sum, na.rm=TRUE)
  }else{
  ctppDataAggreg <- aggregate(list(ctppData[3:12]), by=list(RESIDENCE = ctppData$RESIDENCE), FUN=sum, na.rm=TRUE)
}

ctppDataAggreg

```

## Add new percentage columns for each quantitative field

```{r add_percentages}

for(i in colnames(ctppDataAggreg[-1])){
  ctppDataAggreg[[paste0(i, "Pct")]] <- round((ctppDataAggreg[[i]]/sum(ctppDataAggreg[[i]]))*100, digits = 2)
}

ctppDataAggreg

```


## Load Greater TADs geo layer

```{r load_geo_layer}

# Read the "Greater_TADs" shapefile
shpToRead <- paste(home, boundarySourceFolder, geoLayerName, ".shp", sep="")
geoLayer = st_read(shpToRead)

# Keep only the columns of relevance
geoLayer <- geoLayer[c("neutral_id", "geometry")]

#str(geoLayer)
#geoLayer

# Plot it
plot(st_geometry(geoLayer), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```


## Merge geolayer and table data

```{r merge_geo_and_data}

# Join the non-PHL side to the Greater TADs.
if (PhlSide == "RESIDENCE"){
  geoLayerTads <- merge(geoLayer, ctppDataAggreg, by.x="neutral_id", by.y="WORKPLACE")
}else{
  geoLayerTads <- merge(geoLayer, ctppDataAggreg, by.x="neutral_id", by.y="RESIDENCE")
}

geoLayerTads

```


## Plot resulting TAD-level map

The field Total_Est is displayed.

```{r plot_map, include=TRUE}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

ggplot() + geom_sf(data = geoLayerTads,  aes(fill = Total)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```


## Export to Shp

```{r export_to_shp, include=TRUE}

shpToWrite <- paste(home, outFolder, "MeansOfT_TADS_", flowName, "_full.shp", sep="")
#shpToWrite
st_write(geoLayerTads, shpToWrite, delete_layer=TRUE)

```

# Part II: Generate the County level feature class

## Collapse Greater side to 8 counties

```{r collapse_greater}

# Read the TAD/County crosswalk table 
fileToRead <- paste(home, boundarySourceFolder, "Greater_TADs_and_Counties.xlsx", sep="")
tadsAndCounties <- read_xlsx(fileToRead, col_names = TRUE)

# Join the Greater side (i.e., the non-PHL side) to the TAD/County crosswalk table 
if (PhlSide == "RESIDENCE"){
  ctppDataAggregCo <- merge(ctppDataAggreg, tadsAndCounties, by.x="WORKPLACE", by.y="neutral_id")
}else{
  ctppDataAggregCo <- merge(ctppDataAggreg, tadsAndCounties, by.x="RESIDENCE", by.y="neutral_id")
}

# Collapse the Greater side (i.e., the non-PHL side)
ctppDataAggregCo <- aggregate(list(ctppDataAggregCo[2:11]), by=list(CO_NAME = ctppDataAggregCo$CO_NAME), FUN=sum, na.rm=TRUE)
#ctppDataAggregCo

```

## Add percentage column 

```{r add_percentages_CO}

for(i in colnames(ctppDataAggregCo[-1])){
  #print(i)
  ctppDataAggregCo[[paste0(i, "Pct")]] <- round((ctppDataAggregCo[[i]]/sum(ctppDataAggregCo[[i]]))*100, digits = 2)
}

ctppDataAggregCo

max(ctppDataAggregCo$Total)

ctppDataAggregCo[order(-ctppDataAggregCo$Total),]

ctppDataAggreg[ctppDataAggreg$Res_Area == "Upper North East",]

PhlSide


```

## Create Greater Counties geo layer

```{r create_geo_layer2}

# Read the "Greater_Counties" shapefile
layerName <- "Greater_Counties"
shpToRead <- paste(home, boundarySourceFolder, layerName, ".shp", sep="")
geoLayerCo = st_read(shpToRead)

# Keep only the columns of relevance
geoLayerCo <- geoLayerCo[c("FIPS", "CO_NAME", "geometry")]

#str(geoLayerCo)
geoLayerCo

# Plot it
plot(st_geometry(geoLayerCo), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)


```

## Merge geolayer and data

```{r merge_geo_and_data}

geoLayerCo <- merge(geoLayerCo, ctppDataAggregCo, by.x="CO_NAME", by.y="CO_NAME")
geoLayerCo

```

## Plot resulting County-level map

```{r plot_map2}

ggplot() + geom_sf(data = geoLayerCo,  aes(fill = Total)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```


## Export to Shp

```{r export_to_shp}

shpToWrite <- paste(home, outFolder,  "MeansOfT_COs_", flowName, "_full.shp", sep="")
st_write(geoLayerCo, shpToWrite, delete_layer=TRUE)

```

