---
title: "Gender_by_MeansOfT_PHL"
---

# Summary

* Process the data downloaded from the Census fact finder: Gender by Means Transportation for Philadelphia County as a residence (B08006, ACS 2016 5 years). 
* Input: Single CSV file representing the means of transportation breakdown by gender (3 rows and 105 columns).  

* Output: Produces an Excel file where the data has been restructured and aggregated (5 columns and 51 rows). The rest of the data formatting was done by hand.

* Last modified: 11/07/2019

# Init

Initialize the paths and names. 

```{r init}

library(zoo)
library(dplyr)
library(writexl) 
library(sf)
library(ggplot2)

home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data3rdPass/"

tableName <- "GenderByModeOfT_PHL/ACS_16_5YR_B08006"

tableSourceFolder <- "GisReadyTables/"
outFolder <- "DisplayReadyData/"

```

## Read the raw CSV table and transpose it

```{r read_csv}

# Read the file
fileToRead <- paste(home, "RawData/RawXls/", tableName, "_with_ann.csv", sep="")
fileToRead
ctppData <- read.csv(file=fileToRead, header=FALSE, sep="," ,skip=0)

# Transpose rows into columns to get a dataframe of 4 columns and 573 rows
ctppData <- data.frame(t(ctppData))

# Rename the columns
colnames(ctppData) <- c("ID", "Pretty_Name", "Num_of_People")

# Remove the first 3 rows, which contain geography IDs
ctppData <- ctppData[ -c(1, 2, 3), ]

#ctppData <- ctppData2

#ctppData

```

## Do some cleanup

```{r cleanup_ids}

# Keep only the rows containing Estimates (drop the Margin of Error rows)
ctppData <- subset(ctppData, grepl("Estimate", Pretty_Name),)

# Create new column Gender
ctppData$Gender <- ctppData$Pretty_Name
ctppData$Gender <- sub("^.*Female:.*$", "Female", ctppData$Gender)
ctppData$Gender <- sub("^.*Male:.*$", "Male", ctppData$Gender)
ctppData$Gender <- sub("^.*Total:.*$", "Total", ctppData$Gender)

# Create new column MeansOfT_Category
ctppData$MeansOfT_Category <- ctppData$Pretty_Name
ctppData$MeansOfT_Category <- sub("Estimate; Total: - \\S*: - ", "", 
                                   ctppData$MeansOfT_Category)
ctppData$MeansOfT_Category <- sub("Estimate; Total: - ", "", ctppData$MeansOfT_Category)
ctppData$MeansOfT_Category <- sub("Estimate; Total:", "Total:", ctppData$MeansOfT_Category)
ctppData$MeansOfT_Category <- sub(":$", "", ctppData$MeansOfT_Category)

# Handle the special case of Male and Female total rows
ctppData$MeansOfT_Category[ctppData$MeansOfT_Category == "Male"] <- "Total"
ctppData$MeansOfT_Category[ctppData$MeansOfT_Category == "Female"] <- "Total"

ctppData <- ctppData[c("Gender","MeansOfT_Category", "Num_of_People")] 

ctppData

```


## Convert quantitative cells from text columns to numeric

``` {r text_to_numeric}

ctppData$Num_of_People <- as.character(ctppData$Num_of_People)
ctppData$Num_of_People <- as.numeric(ctppData$Num_of_People)

ctppData
```

## Generate a dataframe with one column per Gender

```{r one_column_per_bracket}

# Create a dataframe seed with just a MeansOfT_Category column
merged_df <- ctppData[ctppData$Gender == "Total",]
merged_df <- merged_df[2] 

# Get the list of all unique income brackets in the original data
Gender_List  <- unique(ctppData$Gender)

# Loop through the list of Genders to get the number of people values for each Gender in each MeansOfT category
# and add them as a new column to the new dataframe  
for(i in Gender_List){
  ctppDataTemp <- ctppData[ctppData$Gender == i,]
  colnames(ctppDataTemp)[colnames(ctppDataTemp)=="Num_of_People"] <- i
  ctppDataTemp <- ctppDataTemp[c(2,3)]
  merged_df <- merge(merged_df, ctppDataTemp, by="MeansOfT_Category")
}

# Sum up Male and Female (just to check)
merged_df$Sum_to_check = merged_df$Male + merged_df$Female

merged_df

```

## Transpose rows into columns again

[NOT USED]

```{r transpose_again}

# Transpose rows into columns to get a dataframe of 11 columns and 27 rows 
ctppDataAggreg <- data.frame(t(merged_df[-1]))
colnames(ctppDataAggreg) <- merged_df[, 1]

ctppDataAggreg
```

## Aggregate some of the means of transportation columns

[NOT USED]

``` {r aggregate_cols eval=FALSE}

# This is not implemented, as it was simpler to do it in the spreadsheet, but as an example:

# Sum up Bicycle and Walked into "Active"
ctppDataAggreg["Active"] <-rowSums(ctppDataAggreg[,c("Bicycle", "Walked")])

ctppDataAggreg

```

## Export to Excel

```{r export_to_excel}

fileToWrite <- paste(home, outFolder, tableName, "_PREPPED.xlsx", sep="")
write_xlsx(x = ctppDataAggreg, path = fileToWrite, col_names = TRUE)

```

