---
title: "Income_by_MeansOfT_PHL"
---

# Summary

* Process the CTPP data: Income by Means Transportation for Philadelphia County as a residence. 

* Input: Single CSV file extracted from the CTPP site representing the means of transportation breakdown for 25 different income brackets (4 rows and 573 columns).  

* Output: Produces an Excel file where the data has been restructured and aggregated (8 columns and 6 rows). It is ready to generate pie or bar charts.

* Last modified: 11/05/2019

# Init

Initialize the paths and names. 

```{r init}

library(zoo)
library(dplyr)
library(writexl) 
library(sf)
library(ggplot2)

home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data3rdPass/"

tableName <- "IncomeByModeOfT_PHL"

tableSourceFolder <- "GisReadyTables/"
outFolder <- "DisplayReadyData/"

```

## Read the raw CSV table and transpose it

```{r read_csv}

# Read the file
fileToRead <- paste(home, "RawData/RawXls/", tableName, ".csv", sep="")
fileToRead
ctppData <- read.csv(file=fileToRead, header=FALSE, sep="," ,skip=3)

# Remove that last line of the table (which is a note)
ctppData <- head(ctppData,-1)

# Transpose rows into columns to get a dataframe of 4 columns and 573 rows
ctppData2 <- data.frame(t(ctppData[-1]))
colnames(ctppData2) <- ctppData[, 1]

# Remove the Residence column, which is empty
ctppData3 <- ctppData2[c(1,2,3,5)]

# Rename the columns
colnames(ctppData3) <- c("Income_bracket", "MeansOfT_Category", "Value_Type", "Num_Of_People")

ctppData <- ctppData3

ctppData

```

## Do some cleanup

```{r cleanup_ids}

# Fill up empty Income_Bracket cells (due to the "cell merge" presentation) with last previous value
ctppData$Income_bracket[ctppData$Income_bracket == ""] <- NA
ctppData$Income_bracket <- na.locf(ctppData$Income_bracket)

# Keep only the rows containing Estimates (drop the Margin of Error rows)
ctppData <- ctppData[ which(ctppData$Value_Type == 'Estimate'),]

# Remove column Value_Type which is now useless
ctppData <- ctppData[c(1,2,4)]

ctppData

```


## Convert quantitative cells from text columns to numeric

``` {r text_to_numeric}

ctppData$Num_Of_People <- as.numeric(gsub(",", "", ctppData$Num_Of_People))

ctppData
```

## Generate a dataframe with one column per income bracket

```{r one_column_per_bracket}

# Create a dataframe seed with just a MeansOfT_Category column
merged_df <- ctppData[ctppData$Income_bracket == "Total, household income",]
merged_df <- merged_df[2] 

# Get the list of all unique income brackets in the original data
income_category  <- unique(ctppData$Income_bracket)

# Loop through the list of income brackets to get the number of people values for each bracket in each MeansOfT category
# and add them as a new column to the new dataframe  
for(i in income_category){
  ctppDataTemp <- ctppData[ctppData$Income_bracket == i,]
  colnames(ctppDataTemp)[colnames(ctppDataTemp)=="Num_Of_People"] <- i
  ctppDataTemp <- ctppDataTemp[c(2,3)]
  merged_df <- merge(merged_df, ctppDataTemp, by="MeansOfT_Category")
}

merged_df

```

## Transpose rows into columns again

```{r transpose_again}

# Transpose rows into columns to get a dataframe of 11 columns and 27 rows 
ctppDataAggreg <- data.frame(t(merged_df[-1]))
colnames(ctppDataAggreg) <- merged_df[, 1]

ctppDataAggreg
```

## Aggregate some of the means of transportation columns

``` {r aggregate_cols}

# Sum up the 3 car-related categories into "Auto"
ctppDataAggreg["Auto"] <-rowSums(ctppDataAggreg[,c("Car, truck, or van -- Drove alone", "Car, truck, or van -- In a 2-person carpool", "Car, truck, or van -- In a 3-or-more person carpool")])

# Sum up all public transit into "Transit"
ctppDataAggreg["Transit"] <-rowSums(ctppDataAggreg[,c("Bus or trolley bus", "Railroad or ferryboat",	"Streetcar, trolley car, subway, or elevated")])

# Sum up Bicycle and Walked into "Active"
ctppDataAggreg["Active"] <-rowSums(ctppDataAggreg[,c("Bicycle", "Walked")])

# Sum up all others into "Other"
ctppDataAggreg["Other"] <-rowSums(ctppDataAggreg[,c("Taxicab, motorcycle or other method", "Worked at home")])

# Remove detail columns
ctppDataAggreg <- ctppDataAggreg[c("Bus or trolley bus", "Railroad or ferryboat", "Streetcar, trolley car, subway, or elevated",  "Auto", "Transit", "Active", "Other", "Total, means of transportation")]

colnames(ctppDataAggreg) <- c("Bus or trolley", "Railroad", "Subway",  "Auto", "Transit", "Active", "Other", "Total")
  
ctppDataAggreg

```

## Export to Excel

```{r export_to_excel}

fileToWrite <- paste(home, outFolder, tableName, "PREPPED.xlsx", sep="")
write_xlsx(x = ctppDataAggreg, path = fileToWrite, col_names = TRUE)

```

