---
title: "Poverty_Status_TADlevel_10Counties_ImprovedAccuracy"
---

# Summary

* Process the CTPP data: Poverty Status at the TAD level for 10 counties (9 DVRPC counties + New Castle, DE.). The goal is to get numbers that are more accurate than the original TAD ones by using other data sources.

* Input: Several CVS files extracted from the CTPP site.  

* Output: Produces a feature class at the County level with 6 data colums: (1) where do residents of philadelphia work (All, Under poverty line, Above poverty line)? and (2) where do workers in philadelphia come from  (All, Under poverty line, Above poverty line)? Both are merged and exported into the same Shp. The same data is also exported to Excel.


* Last modified: 09/26/2019


# Init

Initialize the paths and names. 

```{r init}

library(zoo)
library(dplyr)
library(writexl) 
library(sf)
library(ggplot2)
library(readxl) 


home <- "U:/FY2020/Transportation/TransitBikePed/CityTransitPlan/Delphine_CityTransitPlan/CtppAnalysis/Data2ndPass/"

theme <- "PovertyStatus"
category1 <- "_Munlevel_PHL_to_Suburbs"
category2 <- "_Munlevel_Suburbs_to_PHL"
category3 <- "_TadLevel_PHL_to_NewCastle"
category4 <- "_TadLevel_NewCastle_to_PHL"
category5 <- "_TadLevel_PHL_to_PHL"

tableSourceFolder <- "RawData/RawXls/improved_tads/"
boundarySourceFolder <- "RelevantBoundaries/"
outFolder <- "DisplayReadyData/"

geoLayerName <- "TADlevel_10counties"


```


# Part I: Read the table data and clean it up

## Read the raw CSV table

```{r read_csv}

load_file <- function(category){
  # Read the file
  fileToRead <- paste(home, tableSourceFolder, theme, category, ".csv", sep="")
  fileToRead
  ctppData <- read.csv(file=fileToRead, header=FALSE, sep="," ,skip=6)

  # Remove that last line of the table (which is a note)
  ctppData <- head(ctppData,-1)
  
  # Remove the Margin of Error fields
  noMoE <- c(1,2,3,5,7,9)
    
  ctppData <- ctppData[noMoE]
  
  # Rename the fields
  names(ctppData) <- c("RESIDENCE",	"WORKPLACE",	"Total", "Under_PL", 
                       "Above_PL_100_to_149", "Above_PL_150_and_more")

  #str(ctppData)
  ctppData
}

PHLToSuburbs <- load_file(category1)
SuburbsToPHL <- load_file(category2)
PHLToNewCastle <- load_file(category3)
NewCastleToPHL <- load_file(category4)
PHLToPHL <- load_file(category5)


PHLToSuburbs
SuburbsToPHL
PHLToNewCastle
NewCastleToPHL
PHLToPHL

```

## Clean up

```{r cleanup}

cleanup <- function(ctppData){
  #  Clean up ID columns: Fill up empty Residence cells (due to the "cell merge" presentation) with 
  #  last previous value
  ctppData$RESIDENCE[ctppData$RESIDENCE == ""] <- NA
  ctppData$RESIDENCE <- na.locf(ctppData$RESIDENCE)
  
  # Find the number of columns in the dataframe
  numCols <- ncol(ctppData)
  
  # Turn all quantitative cells to the numeric type, after having removed the potential comma. 
  ctppData[3:numCols]<- lapply(ctppData[3:numCols], function(y) as.numeric(gsub(",", "", y)))
  
  ctppData
}

# Create TAD IDs from the human-readable strings provided (for natively TAD-level data only)
cleanupIds <- function(ctppData){
  # Create IDs out of Residence and Workplace strings
  ctppData$RESIDENCE <- gsub('TAD (\\d+), MPO (\\d+)', 'C0200US\\2\\1', ctppData$RESIDENCE)
  ctppData$WORKPLACE <- gsub('TAD (\\d+), MPO (\\d+)', 'C0200US\\2\\1', ctppData$WORKPLACE)
  
  ctppData
}

PHLToSuburbs <- cleanup(PHLToSuburbs)
SuburbsToPHL <- cleanup(SuburbsToPHL)

PHLToNewCastle <- cleanup(PHLToNewCastle)
PHLToNewCastle <- cleanupIds(PHLToNewCastle)

NewCastleToPHL <- cleanup(NewCastleToPHL)
NewCastleToPHL <- cleanupIds(NewCastleToPHL)
PHLToPHL <- cleanup(PHLToPHL)
PHLToPHL <- cleanupIds(PHLToPHL)

PHLToSuburbs
SuburbsToPHL
PHLToNewCastle
NewCastleToPHL
PHLToPHL

```

# Part II: Generate TAD-level feature class

## Aggregate from Municipalities to TADs 
(for municipality-level data only)

```{r aggregate_to_TADs}

aggregatetoTADs <- function(ctppData, residence){
  
  # Read the Municipalities/TADs crosswalk table 
  crossWalkFile <- paste(home, boundarySourceFolder, "Mun_to_TAD_crosswalk.xlsx", sep="")
  munToTADs <- read_xlsx(crossWalkFile, col_names = TRUE)
  
  munToTADs
  
  # Join the Suburbs side (i.e., the non-PHL side) to the Municipalities/TADs crosswalk table
  if (residence == "PHL"){
    ctppDataWithTADs <- merge(ctppData, munToTADs, by.x="WORKPLACE", by.y="name")
    ctppDataWithTADs <- ctppDataWithTADs[c("WORKPLACE", "neutral_id", "Total", "Under_PL", 
                       "Above_PL_100_to_149", "Above_PL_150_and_more")]
  }else{
    ctppDataWithTADs <- merge(ctppData, munToTADs, by.x="RESIDENCE", by.y="name")
    ctppDataWithTADs <- ctppDataWithTADs[c("RESIDENCE", "neutral_id", "Total", "Under_PL", 
                       "Above_PL_100_to_149", "Above_PL_150_and_more")]
  }
  
  # Find the number of columns in the dataframe
  numCols <- ncol(ctppDataWithTADs)
  
  # Turn all quantitative cells to the numeric type, after having removed the potential comma. 
  ctppDataWithTADs[3:numCols]<- lapply(ctppDataWithTADs[3:numCols], 
                                       function(y) as.numeric(gsub(",", "", y)))
  
  # Then group by the data based on the TAD ID
  ctppDataWithTADs <- aggregate(list(ctppDataWithTADs[3:numCols]), 
                                by=list(TAD_ID = ctppDataWithTADs$neutral_id), FUN=sum, na.rm=TRUE)
  ctppDataWithTADs
}

PHLToSuburbs <- aggregatetoTADs(PHLToSuburbs, residence="PHL")
SuburbsToPHL <- aggregatetoTADs(SuburbsToPHL, residence="suburbs")

PHLToSuburbs
SuburbsToPHL

```


## Collapse PHL to 1 single location then multiply by constant factor 
(for natively TAD-level data only)

```{r collapse_PHL}


# Group by the non-PHL side (which results in collapsing the PHL side)
collapsePHL <- function(ctppData, residence){
  # Find the number of columns in the dataframe
  numCols <- ncol(ctppData)

  if (residence == "PHL"){
    ctppDataAggreg <- aggregate(list(ctppData[3:numCols]), by=list(TAD_ID = ctppData$WORKPLACE), FUN=sum, na.rm=TRUE)
  }else{
    ctppDataAggreg <- aggregate(list(ctppData[3:numCols]), by=list(TAD_ID = ctppData$RESIDENCE), FUN=sum, na.rm=TRUE)
  }
  
  ctppDataAggreg
}


# Multiply by the appropriate constant factor to make up for the Census undercounting of workplaces at the TAD level.
# Note: the constant factors were computed separately based on the difference between county numbers and sums of TADs.

MultiplyByFactor <- function(ctppData, rectifyingfactor){
  ctppData[] = lapply(ctppData, FUN = function(x) if (is.numeric(x)) 
    return(round(x * rectifyingfactor, digits = 0)) else return(x))
  ctppData
}  


PHLToNewCastle <- collapsePHL(PHLToNewCastle, residence="PHL")
PHLToNewCastleRectified <- MultiplyByFactor(PHLToNewCastle, 1.23)
NewCastleToPHL <- collapsePHL(NewCastleToPHL, residence="suburbs")
NewCastleToPHLRectified <- MultiplyByFactor(NewCastleToPHL, 1.21)
PHLToPHLRes <- collapsePHL(PHLToPHL, residence="PHL")
PHLToPHLResRectified <- MultiplyByFactor(PHLToPHLRes, 1.21)
PHLToPHLWp <- collapsePHL(PHLToPHL, residence="OtherSide")
PHLToPHLWpRectified <- MultiplyByFactor(PHLToPHLWp, 1.21)

PHLToNewCastle
PHLToNewCastleRectified
NewCastleToPHL
NewCastleToPHLRectified
PHLToPHLRes
PHLToPHLResRectified
PHLToPHLWp
PHLToPHLWpRectified

```

## Aggregate some of the columns

``` {r aggregate_cols}

aggregateCols <- function(ctppData){
  # Sum up the two levels of "above poverty line" into the new field "Above_PL"
  ctppData["Above_PL"] <-rowSums(ctppData[,c("Above_PL_100_to_149", "Above_PL_150_and_more")])
  
  # Remove detail columns
  ctppData <- ctppData[c("TAD_ID",	"Total", "Under_PL", "Above_PL")]
  
  ctppData
}

PHLToSuburbsCompact <- aggregateCols(PHLToSuburbs)
SuburbsToPHLCompact <- aggregateCols(SuburbsToPHL)
PHLToNewCastleCompact <- aggregateCols(PHLToNewCastleRectified)
NewCastleToPHLCompact <- aggregateCols(NewCastleToPHLRectified)
PHLToPHLResCompact <- aggregateCols(PHLToPHLResRectified)
PHLToPHLWpCompact <- aggregateCols(PHLToPHLWpRectified)

PHLToSuburbsCompact
SuburbsToPHLCompact
PHLToNewCastleCompact
NewCastleToPHLCompact
PHLToPHLResCompact
PHLToPHLWpCompact


```

## Get PHL as a residence and PHL as a workplace in the same dataframe (for  PANJ suburbs, New Castle, and Philadelphia)

```{r mergeBothFlows}

mergeBothFlows <- function(PHLToSuburbsFlow, SuburbsToPHLFlow){
  # Rename the fields according to whether they represent workplace or residence data
  names(PHLToSuburbsFlow) <- c("WORKPLACE", "TotalWp", "UndrPLWp", "AbvePLWp")
  names(SuburbsToPHLFlow) <- c("RESIDENCE", "TotalRe", "UndrPLRe", "AbvePLRe")
  
  BothFlows <- merge(PHLToSuburbsFlow, SuburbsToPHLFlow, by.x="WORKPLACE", by.y="RESIDENCE")
  BothFlows <- BothFlows[c("WORKPLACE",	"TotalWp", "TotalRe", "UndrPLWp",
                           "UndrPLRe", "AbvePLWp", "AbvePLRe")]
  names(BothFlows)[names(BothFlows)=="WORKPLACE"] <- "TAD_ID"
  
  BothFlows
}

PANJ_suburbs <- mergeBothFlows(PHLToSuburbsCompact, SuburbsToPHLCompact)
NewCastle <- mergeBothFlows(PHLToNewCastleCompact, NewCastleToPHLCompact)
Philadelphia <- mergeBothFlows(PHLToPHLResCompact, PHLToPHLWpCompact)

PANJ_suburbs
NewCastle 
Philadelphia

```

## Append the dataframes for the 3 TAD regions (PANJ suburbs, New Castle, and Philadelphia)

```{r Append_3_Regions}

allTadRegionsData <- new <- rbind(PANJ_suburbs, NewCastle, Philadelphia)
allTadRegionsData

```


## Load TAD boundaries shapefile for the 10 counties

```{r load_geo_layer}

# Read the TAD shapefile
shpToRead <- paste(home, boundarySourceFolder, geoLayerName, ".shp", sep="")
geoLayer = st_read(shpToRead)

#str(geoLayer)
geoLayer

# Plot it
plot(st_geometry(geoLayer), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```


## Merge (join) TAD boundaries and data table

```{r merge_geo_and_data}

# Join the non-PHL side to the Greater TADs.
GeoTenCountyTotals <- merge(geoLayer, allTadRegionsData, by.x="neutral_id", by.y="TAD_ID")

GeoTenCountyTotals
```


## Plot resulting TAD-level maps

The field TotalRe or TotalWp is displayed.

```{r plot_map, include=TRUE}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

ggplot() + geom_sf(data = GeoTenCountyTotals,  aes(fill = TotalRe)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

ggplot() + geom_sf(data = GeoTenCountyTotals,  aes(fill = TotalWp)) + theme_bw() + ditch_the_axes + scale_fill_gradient(low="#ccccff", high="#0000e6")

```

# Part III: Export
## Export to Shapefile

```{r export_to_shp, include=TRUE}

shpToWrite <- paste0(home, outFolder, theme, "_TADLevel_10Counties.shp")
st_write(GeoTenCountyTotals, shpToWrite, delete_layer=TRUE)

```


## Export to Excel

```{r export_to_excel}

fileToWrite <- paste0(home, outFolder, theme, "_TADLevel_10Countiesxlsx")
write_xlsx(x = allTadRegionsData, path = fileToWrite, col_names = TRUE)

```
